The evolution of the transcriptome of mammalian placentas

### 2016-09-29 manuscript/manuscript.Rnw
start working on discussion

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and the standard Illumina protocol. Paired-end sequencing
was performed using the Illumina Genome Analyzer II with an insert
size of 230 base pairs (bp). Read lengths for species are shown in
Supplemental \cref{tab:ref_genome_aligned_species}.

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
`Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
genes in the core transcriptome.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

@@

The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation. The core placenta transcriptome is also enriched in
genes involved in maintaining the structure of the placenta and
transporting proteins, as would be expected given the function of the
placenta.

>>>>

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Me\-di\-cine using 2.5-5 µg of
total RNA and the standard Illumina protocol. Paired-end sequencing
was performed using the Illumina Genome Analyzer II with an insert
size of 230 base pairs (bp). Read lengths for species are shown in
Supplemental \cref{tab:ref_genome_aligned_species}.

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
`Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

\subsubsection{Annexin Complexes}

Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

\subsubsection{Invasion: EGF and Auxin Y in the Core Placenta}

@@

\subsubsection{Annexin Complexes}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation. The core placenta transcriptome is also enriched in
genes involved in maintaining the structure of the placenta and
transporting proteins, as would be expected given the function of the
placenta.

\subsubsection{Invasion: EGF and Auxin Y in the Core Placenta}
